State of the Practice for Lattice Boltzmann Method Software

### 2022-04-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Edits completed for Section 2

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template, as described in
\citet{SmithEtAl2021}.  For each software package (each column in the template),
we fill-in the rows of the template. This process takes about two hours per
package, with a cap of four hours. The time constraint is necessary so that the
work load is feasible for a team as small as one, given our aim to cap the
measurement phase at 160 person hours \citep{SmithEtAl2021}.  An excerpt of the
template, in spreadsheet form, is shown in
Figure~\ref{measurement_template_image}.

@@

The questions were designed to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. The measures are grouped under headings for
each quality, and one for summary information
(Figure~\ref{measurement_template_image}). The summary section provides general
information, such as the software name, number of developers, etc.  We follow
the definitions given by \citet{gewaltig2012quality} for the software
categories.  Public means software intended for public use.  Private means
software aimed only at a specific group, while the concept category is used for
software written simply to demonstrate algorithms or concepts. The three
categories of development models are: open source, where source code is freely
available under an open source license; free-ware, where a binary or executable
is provided for free; and, commercial, where the user must pay for the software
product.

Several of the qualities use the word ``surface''.  This is to highlight that,
for these qualities in particular, the best that we can do is a shallow measure.
For instance, we do not conduct experiments to measure usability. Instead, we
are looking for an indication that usability was considered by the developers by
looking for cues in the documentation, such as the presence of getting started
instructions, a user manual or documentation of expected user characteristics.

Tools were used to find some of the measurements, such as the number of files,
number of lines of code (LOC), percentage of issues that are closed, etc. The
tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
each software package's GitHub repository for the number of binary files, the
number of added and deleted lines, and the number of commits over varying time
intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
(scc)} was used to measure the number of text based files as well as the number
of total, code, comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environment for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors.

@@

Once we have measured each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}.  To do this, we used the Analytic Hierarchy Process
(AHP), a decision-making technique that is used to compare multiple options by
multiple criteria. AHP was used for comparing and ranking the LBM software
packages using the overall impression quality scores that were gathered in the
measurement template.  AHP performs a pairwise analysis using a matrix and
generates an overall score as well as individual quality scores for each
software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
software based on quality measures.

This project used a tool we developed for conducting the AHP analysis. The tool
includes a sensitivity analysis to ensure that the software package rankings are
appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis we modified the score by 10\% for each package and verified
that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file of the tool includes requirements and usage information.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interviewed developers
using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
focus on specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often bring up exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit four developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meetings were held using
on-line meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Since the LBM domain has a reasonably small scope, we are able to view the
software as constituting a program family.  The concept of a program family is
defined by \citet{parnas1976design} as ``a set of programs whose common
properties are so extensive that it is advantageous to study the common
properties of the programs before analyzing individual members''. Studying the
common properties within a family of related programs is termed a domain
analysis.

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions and assumptions that
differ between family members. Associated with each variability are its
parameters of variation, which summarize the possible values for that
variability, along with their potential binding time.  The binding time is when
the value of the variability is set.  It could be set as specification time,
build time (when the program is being compiled) or run time (when the code is
executing). As an example, for the current assessment, the packages all have the
commonality of using LBM techniques to simulate the motion of fluids.  For
variabilities, the LBM packages can be distinguished by the dimension of
problems they simulate, whether or not (and how) they support parallel
computation, whether or not compressibility can be considered, whether reflexive
boundaries can be employed, whether the simulation handles multi-fluids, whether
the simulations handles turbulence, and whether complex geometries can be
simulated.

To keep the current presentation focused, we only present the results of our
shallow domain analysis. For the LBM domain a table was constructed that
distinguishes the programs by their variabilities.  In research software the
variabilities are often assumptions, as shown in Table~\ref{tbl_features}. The
variabilities are listed as columns in Table~\ref{tbl_features} with the
software listed in alphabetical order. A more complete analysis of the
commonalities and variabilities between LBM packages is provided in
\citet{Michalski2021}.

@@

We partnered with a Domain Expert to vet our list of projects
(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}).  The
Domain Expert is an important member of the state of the practice assessment
team. Pitfalls exist if non-experts attempt to acquire an authoritative list of
software, or try to definitively rank the software. Non-experts have the problem
that they can only rely on information available on-line, which has the
following drawbacks:
\begin{inparaenum}[i)]
  \item the on-line resources could have false or inaccurate information; and,
  \item the on-line resources could leave out relevant information that is so
in-grained with experts that nobody thinks to explicitly record it.
\end{inparaenum}
For the current assessment, our Domain Expert (and paper co-author) is Dr.\
Zahra Keshavarz-Motamed, Assistant Professor of Mechanical Engineering at
McMaster University, Hamilton, Ontario, Canada.

The Domain Expert has an important role with verifying the list of LBM packages.
In advance of the first meeting with the Domain Expert, they were asked to
create a list of top software packages in the domain.  This is done to help the
expert get in the right mind set in advance of the meeting.  Moreover, by doing
the exercise in advance, we avoid the potential pitfall of the expert approving
the discovered list of software without giving it adequate thought.  The Domain
Expert was also asked to vet the collected data and analysis.  In particular,
they were asked to vet the proposed list of software packages and the AHP
ranking.  These interactions were done via virtual meetings.

>>>>

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template, as described in
\citet{SmithEtAl2021}.  For each software package (each column in the template),
we fill in the rows of the template. This process takes about two hours per
package, with a cap of four hours. The time constraint is necessary so that the
workload is feasible for a team as small as one, given our aim to cap the
measurement phase at 160 person-hours \citep{SmithEtAl2021}.  An excerpt of the
template, in spreadsheet form, is shown in
Figure~\ref{measurement_template_image}.

@@

We designed the questions to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. We categorize them by quality, with an extra
category for summary information (Figure~\ref{measurement_template_image}). The
summary section provides general data, such as the software name, number of
developers, etc. \citet{gewaltig2012quality} provide us with our software
categories. Public means software available to all, while private means software
available only to a specific group.  The concept category defines software
written to demonstrate algorithms or concepts. The three categories of
development models are open source, where source code is freely available under
an open-source license; free-ware, where a binary or executable is provided for
free; and commercial, where the user must pay.

Several of the qualities use the word ``surface'' to highlight that, for these
qualities, the best that we can do is a shallow measure. For instance, we do not
conduct experiments to measure usability directly. Instead, we look for an
indication that usability is considered by searching for cues in the
documentation, such as the presence of getting started instructions, a user
manual or documentation of expected user characteristics.

We used tools to automatically measure the number of files, number of lines of
code (LOC), percentage of closed issues, etc. GitStats measures each package's
GitHub repository for the number of binary files, the number of added and
deleted lines, and the number of commits over varying time intervals. Sloc Cloc
and Code (scc) measures the number of text files and the number of total, code,
comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, we used Virtual machines (VMs) to provide an
optimal testing environment for each package. We used VMs to create fresh
environments, eliminating worry about existing libraries and conflicts.
Moreover, when the tests are complete, the VM can be deleted, without impacting
the host operating system. The most significant advantage of using VMs is to
level the playing field. Every software install starts from a clean slate, which
removes ``works-on-my-computer'' errors.

@@

Once we measure each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP),
a decision-making technique that compares multiple options by multiple criteria.
Our AHP algorithm uses the overall impression score, gathered via the
measurement template, to compare and rank the LBM software. AHP performs a
pairwise analysis using a matrix and generates an individual quality scores for
each software package. \citet{SmithEtAl2016} shows the details on how AHP ranks
software based on quality measures.

Our ranking uses a tool we developed for conducting the AHP analysis. The tool
includes a sensitivity analysis to ensure that the software package rankings are
appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis, we modified the score by 10\% for each package and
verified that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file for the tool includes requirements and usage information.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain insight, we interviewed developers using a list of
20 questions from \citet{SmithEtAl2021}. The questions cover the background of
the development teams, the interviewees, and the software itself. We ask the
developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
focus on specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often raise exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since we added Munubi after our ethics approval expired. We found the developers
from project websites, code repositories, publications, and biographic
information on institution webpages. Four developers agreed to participate in
our study. We met with them online, using either Zoom or Teams to facilitate
recording and automatic transcription. Each meeting lasted about an hour. The
McMaster University Research Ethics Board approved the interview process under
application number
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Since the LBM domain has a reasonably small scope, we can view the software as
constituting a program family.  \citet{parnas1976design} defines a program
family as ``a set of programs whose common properties are so extensive that it
is advantageous to study the common properties of the programs before analyzing
individual members.'' Studying the common properties within a family of related
programs is termed a domain analysis.

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions and assumptions that
differ between family members. Associated with each variability are parameters
of variation, which summarize the possible values for that variability, along
with their potential binding time. The binding time is when the value of a
variability is set. Variabilities are set at specification time, build time
(when the program compiles) or run time (when the code executes). For the
current assessment, the packages all have the commonality of using LBM
techniques to simulate the motion of fluids. Variabilities for LBM packages
include the following: the dimension of problems they simulate, whether or not
(and how) they support parallel computation, whether or not compressibility can
be considered, whether reflexive boundaries can be employed, whether the
simulation handles multi-fluids, whether the simulations handles turbulence, and
whether complex geometries can be simulated.

To keep the current presentation focused, we only present the results of our
shallow domain analysis. For the LBM domain, a table was constructed that
distinguishes the programs by their variabilities. In research software, the
variabilities are often assumptions. Table~\ref{tbl_features} lists the
variabilities as columns, with the software alphabetically ordered.
\citet{Michalski2021} provides a detailed analysis of the commonalities and
variabilities between LBM packages.

@@

We partnered with a Domain Expert to vet our list of projects
(\rqref{RQ_WhatProjects}) and our ranking (\rqref{RQ_HighestQuality}). The
Domain Expert is a crucial member of the state of the practice assessment team.
Pitfalls exist if non-experts attempt to acquire an authoritative list of
software or try to rank the software. Non-experts rely on online information,
which has the following drawbacks:
\begin{inparaenum}[i)]
  \item the online resources could have false or inaccurate information, and 
  \item the online resources could leave out relevant information that is too
  ingrained with experts for them to explicitly record it.
\end{inparaenum}
For the current assessment, our Domain Expert (and paper co-author) is Dr.\
Zahra Keshavarz-Motamed, Assistant Professor of Mechanical Engineering at
McMaster University, Hamilton, Ontario, Canada.

The Domain Expert has a central role in verifying the list of LBM packages. In
advance of the first meeting with the Domain Expert, we asked them to create a
list of top software packages in the domain.  We did this to help the expert get
in the right mindset for our meeting.  Moreover, by doing the exercise in
advance, we avoid the potential pitfall of the expert approving the discovered
list of software without giving it adequate thought. We asked the Domain Expert,
via virtual meetings, to vet the collected data and analysis. In particular, we
asked them to vet the proposed list of software packages and the AHP ranking.
